### 2021-03-11 manuscript.Rmd
dca added

---
title: "Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation"
---

## Introduction

A capnography waveform displays the level of expired carbon dioxide (CO~2~) over time to show changes in concentrations throughout the respiratory cycle. Capnography waveform abnormalities assist in the detection and diagnosis of specific conditions, such as partial airway obstruction and apnea. For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities including peak professional organisations for anaesthesia in Canada, the United States and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018]

A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue. Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation. presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds. It would be ideal for clinicians to receive capnography monitor alarms as *_early as possible_* in the course of an apneic event, *but only* if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.

It is possible that  capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians earlier for apneic events that are predicted to be prolonged. Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.

Based on results from prior research, it seems that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods. It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states. Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department. It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).

We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach. The aim was to determine the accuracy of machine learning models for predicting if a period of apnea will persist for 30 seconds or more.

# Methods

A secondary analysis of a prospective observational study was undertaken. Results for the primary aim of the observational study are reported elsewhere.[@conway2019sequence] All participants provided written informed consent and the study was approved by human research ethics committees (UCH HREC 1614; SVHAC HREC 16/26; QUT 1600000641). The study was undertaken in accordance with the Australian National Statement on Ethical Conduct in Human Research[@national2007] and it was registered prospectively (ACTRN12616001132437).

### Prediction Goal

The prediction goal was to classify apneic events at the 15-second time-point as being either short (i.e. terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer). The prediction algorithm was compared against a default alarm setting for capnography monitors, which triggers an alarm at 30 seconds of apnea.

### Participants

Participants in the study were consecutive adult patients scheduled to undergo an elective procedure in the cardiac catheterization laboratory with moderate sedation. Patients with severe cognitive impairment (due to inability to provide informed consent) or inability to understand and speak English (if an interpreter was unavailable) were excluded.

### Sedation and monitoring

The sedation regimen used for patients included in this study comprised bolus doses of intravenous midazolam and fentanyl. Sedation was administered by nurses who were trained in advanced life support. Routine clinical monitoring included cardiac rhythm and pulse oximetry monitoring as well as non-invasive blood pressure measurements every 5-10 minutes. The Respironics LoFlo sidestream CO~2~ sensor was used for capnography monitoring. A CO~2~ sampling cannula was inserted into the sideport of an oxygen face-mask or was integrated as a separate line for nasal cannulas. The capnography waveform was displayed on the main physiological monitoring screen. A default ‘No breaths detected’ alert was triggered for apnea but no other audible or visual alarms were set for the capnography monitor. No restrictions or specific instructions were provided to clinicians regarding the detection of capnography waveform abnormalities as part of the research protocol because the study used observational design.

### Data collection

Demographic data and clinical characteristics were collected from medical records and self-report questionnaires prior to procedures. Intra-procedural data were collected in real-time by the researcher who was present in the procedure room. Direct observation of the participant was required to record the timing of sedation administrations and any interventions applied by sedation providers.

### Predictor variables

### Statistical analysis

The best model was chosen with respect to the optimal balance between maximising the ability to predict prolonged events and alarm frequency. The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds). A true positive was an apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds. A false positive was an apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds.

#### Modelling

#### Decision curve analysis

We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than not intervening for patients who had apnea for >15 seconds (i.e. the default strategy). The net benefit is calculated using the formula:

This formula essentially transforms the total number of true positives and false positives into a standardized scale, being the number of true positives, weighted according to the probability threshold. For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model. This net benefit can result from any combination of true positives and false positives.

A decision curve was created to plot net benefits across a range of probability thresholds, which take into account a plausbile range of clinician preferences for the point at which an intervention should be initiated to treat apnea (this could be, for example, something as simple as verbal stimulation). We tested the thresholds in the range from 0.4 to 0.8. In a practical sense, this means that we decided that all clinicians would not ever accept a probability for prolonged apnea lower than 0.4 as a useful trigger for intervention, and, conversely, that all clinicians would always consider that intervention is indicated if the probability of prolonged apnea was higher than 0.8. These probability thresholds can be interpreted as value preferences in the context of decision curve analysis. For example, a clinician who is 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4). Individual participant characteristics will also influence clinicians' decisions about probability thresholds. A clinician may elect to intervene when the probability for prolonged apnea is 0.5 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.

A threshold of 0.5 inherently indicates that avoiding a false positive is as valuable of an outcome as identifiying a true positive. Preferences for probability thresholds below 0.5 are weighted such that identifying a true positive is more valuable than avoiding a false positive. Preferences for probability thresholds above 0.5 are weighted such that avoiding a false positive is more valuable than identifying a true positive.  Decision curves can be interpreted such that the strategy with the highest net benefit at each probability threshold has the highest clinical value.[@vickers2016net]

## Results

### Decision curve analysis

Net benefit associated with the random forest model exceeded the strategy of intervening for all or no patients across all probability thresholds from the range of 0.4 to 0.8. The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was 40%, would be achieved by using the random forest model as opposed to treating everyone or no one.o

### 2021-03-11 R/dca.r

@@ -1,0 +1,213 @@
+#' Decision curve analysis
+#' @export
+#' 
+dca <- function(data, outcome, predictors, xstart=0.01, xstop=0.99, xby=0.01, 
+  ymin=-0.05, probability=NULL, harm=NULL,graph=TRUE, intervention=FALSE, 
+  interventionper=100, smooth=FALSE,loess.span=0.10) {
+  
+
+  # data MUST BE A DATA FRAME
+  if (class(data)!="data.frame") {
+      stop("Input data must be class data.frame")
+  }
+  
+  #ONLY KEEPING COMPLETE CASES
+  data=data[complete.cases(data[append(outcome,predictors)]),append(outcome,predictors)]
+
+  # outcome MUST BE CODED AS 0 AND 1
+  if (max(data[[outcome]])>1 | min(data[[outcome]])<0) {
+    stop("outcome cannot be less than 0 or greater than 1")
+  }
+  # xstart IS BETWEEN 0 AND 1
+  if (xstart<0 | xstart>1) {
+    stop("xstart must lie between 0 and 1")
+  }
+  
+  # xstop IS BETWEEN 0 AND 1
+  if (xstop<0 | xstop>1) {
+    stop("xstop must lie between 0 and 1")
+  }
+  
+  # xby IS BETWEEN 0 AND 1
+  if (xby<=0 | xby>=1) {
+    stop("xby must lie between 0 and 1")
+  }
+  
+  # xstart IS BEFORE xstop
+  if (xstart>=xstop) {
+    stop("xstop must be larger than xstart")
+  }
+     
+  #STORING THE NUMBER OF PREDICTORS SPECIFIED
+  pred.n=length(predictors)
+  
+  #IF probability SPECIFIED ENSURING THAT EACH PREDICTOR IS INDICATED AS A YES OR NO
+  if (length(probability)>0 & pred.n!=length(probability)) {
+    stop("Number of probabilities specified must be the same as the number of predictors being checked.")
+  }
+ 
+  #IF harm SPECIFIED ENSURING THAT EACH PREDICTOR HAS A SPECIFIED HARM
+  if (length(harm)>0 & pred.n!=length(harm)) {
+    stop("Number of harms specified must be the same as the number of predictors being checked.")
+  }
+  
+  #INITIALIZING DEFAULT VALUES FOR PROBABILITES AND HARMS IF NOT SPECIFIED
+  if (length(harm)==0) {
[…]

### 2021-03-11 _targets.R

@@ -1,12 +1,10 @@
 library(targets)
-library(tidymodels)
-library(tidyverse)
-
+library(tarchetypes)
 
 devtools::load_all()
+options(tidyverse.quiet = TRUE)
 
-
-
+tar_option_set(packages = c("tidymodels", "tidyverse"))
 list(
   # Load in patient data and respiratory state data
   # and reformat into dataframe
@@ -68,13 +66,12 @@ list(
       all_nominal(),
       -has_role("id variable"),
       -all_outcomes()
-    )
-    ),
+    )),
 
   tar_target(wf, workflow() %>%
     add_recipe(base_rec)),
 
-  tar_target(class_metrics, metric_set(roc_auc, sens)),
+  tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
 
   # random forest model
   tar_target(rf, rand_forest(trees = 50) %>%
@@ -86,7 +83,7 @@ list(
       add_model(rf) %>%
       fit_resamples(folds,
         # metrics = class_metrics,
-        control = control_resamples(save_pred = TRUE) # add this line
+        control = control_resamples(save_pred = TRUE)
       )
   ),
   tar_target(rf_metrics, collect_metrics(rf_fit)),
@@ -98,13 +95,50 @@ list(
     ppv(estimate = factor(pred), truth = long) %>%
     pull(.estimate)),
 
+  tar_target(rf_sens, rf_preds %>%
+    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
+    sensitivity(estimate = factor(pred), truth = long) %>%
+    pull(.estimate)),
+
+  tar_target(rf_f_meas, rf_preds %>%
+    mutate(pred = ifelse(.pred_long > 0.3, "long", "short")) %>%
+    f_meas(estimate = factor(pred), truth = long) %>%
+    pull(.estimate)),
+  tar_target(rf_precision, rf_preds %>%
+    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
+    precision(estimate = factor(pred), truth = long) %>%
+    pull(.estimate)),
+  tar_target(rf_recall, rf_preds %>%
+    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
+    recall(estimate = factor(pred), truth = long) %>%
+    pull(.estimate)),
+
+
+
   # Only when optimized feature selection and tuning
[…]

### 2021-03-11 scratch.R

@@ -36,3 +36,40 @@ filter(pred=="short" & long =="long")
         filter(dur >= lower_trigger,
         ID %in% resp_data$ID) %>% View()
 
+data <- testing %>%
+ mutate(outcome = ifelse(long == "long", 1, 0))
+
+ data$rf <- rf_test$.pred_long
+
+test <- dca(data = as.data.frame(data),
+        outcome = "outcome",
+        predictors = "Random forest",
+        smooth = TRUE,
+        xstop = 0.8
+      )
+
+test$net.benefit %>%
+select(-rf_sm,
+"Threshold" = threshold,
+"All" = all) %>%
+pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
+ggplot(aes(x = Threshold, y= `Net benefit`)) +
+geom_line(aes(color = Decisions)) +
+coord_cartesian( ylim = c(-0.1, 0.5))
+
+
+
+rf_test %>%
+    mutate(pred = ifelse(.pred_long > 0.4, "long", "short")) %>%
+    mutate(tp = ifelse(pred == "long" & long == "long", 1, 0)) %>%
+    mutate(fp = ifelse(pred == "long" & long == "short", 1, 0)) %>%
+    summarize(tpos = sum(tp)/nrow(rf_test),
+    fpos = sum(fp)/nrow(rf_test),
+    nb = tpos - fpos * (0.4/0.6))
+
+
+
+    ppv(estimate = factor(pred), truth = long) %>%
+    pull(.estimate)
+
+

### 2021-03-12 manuscript.Rmd
combined plot

@@
 Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation.
-presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds.
+Presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds.
 It would be ideal for clinicians to receive capnography monitor alarms as *_early as possible_* in the course of an apneic event, *but only* if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.
@@
 
-Based on results from prior research, it seems that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods.
-It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
-Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
-It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).
-
-We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.
 The aim was to determine the accuracy of machine learning models for predicting if a period of apnea will persist for 30 seconds or more.
@@
 
-Net benefit associated with the random forest model exceeded the strategy of intervening for all or no patients across all probability thresholds from the range of 0.4 to 0.8.
-The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was 40%, would be achieved by using the random forest model as opposed to treating everyone or no one.o
+Net benefit associated with the random forest model exceeded the strategy of intervening for all patients across all probability thresholds from the range of 0.4 to 0.8.
+The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, would be achieved by using the random forest model as opposed to treating everyone.
+In contrast, the net benefit associated with the random forest model was lower than the strategy of not intervening for any patients across all probability thresholds from the range of 0.4 to 0.8.
+This was due to the fact that we applied a condition that we would not accept more than 5 false positives to detect 1 true positive.
+The number of false positives that triggered across probability thresholds from 0.4 to 0.8 using the random forest model would far exceed the number of true positives, meaning the preferred strategy should remain as only setting the alert for the 30-second time-point.
+
+## Discussion
+
+Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods.
+It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
+Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
+It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).
+
+We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.
+
+Using the machine learning model would produce a better clinical outcome, in terms of the balance between true positives and false positives for detection of apneic periods over 30 seconds, than a default strategy of intervening for everyone who has an apneic period ≥15 seconds.

### 2021-03-12 _targets.R

@@ -4,7 +4,11 @@ library(tarchetypes)
 devtools::load_all()
 options(tidyverse.quiet = TRUE)
 
-tar_option_set(packages = c("tidymodels", "tidyverse"))
+tar_option_set(packages = c(
+  "tidymodels",
+  "tidyverse",
+  "patchwork"
+))
 list(
   # Load in patient data and respiratory state data
   # and reformat into dataframe
@@ -74,7 +78,7 @@ list(
   tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
 
   # random forest model
-  tar_target(rf, rand_forest(trees = 50) %>%
+  tar_target(rf, rand_forest(trees = 500) %>%
     set_engine("randomForest") %>%
     set_mode("classification")),
   tar_target(
@@ -186,7 +190,7 @@ list(
       `Random forest` = rf_test$.pred_long
     )),
 
-  tar_target(dca, dca(
+  tar_target(dca_none, dca(
     data = as.data.frame(dca_data),
     outcome = "outcome",
     predictors = "Random forest",
@@ -194,20 +198,61 @@ list(
     xstart = 0.4,
     xstop = 0.8,
     graph = F,
-    # harm = 1/5 # wouldn't accept needing to intervene on 5 false postitives to get a true positive
+    # wouldn't accept needing to intervene on 5 false postitives to get a true positive
+    harm = 1 / 5
   )),
 
-  tar_target(dca_plot, dca$net.benefit %>%
+  tar_target(dca_plot_none, dca_none$net.benefit %>%
     select(-`Random forest_sm`,
       "Threshold" = threshold,
       -all,
-      "None" = none
+      "Alarm at ≥30 seconds" = none
     ) %>%
-    pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
+    pivot_longer(
+      cols = -Threshold,
+      names_to = "Decisions",
+      values_to = "Net benefit"
+    ) %>%
     ggplot(aes(x = Threshold, y = `Net benefit`)) +
     geom_line(aes(linetype = Decisions)) +
-    coord_cartesian(ylim = c(-0.2, 0.5)) +
+    coord_cartesian(ylim = c(-0.2, 0.2)) +
     theme_minimal() +
-    theme(legend.position = "bottom")
-    ) 
+    theme(
+      legend.position = "bottom",
+      legend.title = element_blank()
[…]

### 2021-03-12 manuscript.Rmd
dca - no harm

@@
 title: "Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation"
+bibliography: "Extras/references.bib"
 ---
@@
 Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation.
-Presumably in an attempt to reduce alarm burden, capnography monitor alarms now mostly default to triggering after apnea has been observed for 30 seconds.
-It would be ideal for clinicians to receive capnography monitor alarms as *_early as possible_* in the course of an apneic event, *but only* if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.
+In practice, two alternative strategies for capnography alarm management are typically used.
+The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (15 seconds).
+The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (30 seconds).
 
-It is possible that  capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians earlier for apneic events that are predicted to be prolonged.
+It is possible that capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
 Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.
 
-The aim was to determine the accuracy of machine learning models for predicting if a period of apnea will persist for 30 seconds or more.
+For the aggressive strategy, only triggering an alarm for apneas predicted to be prolonged would reduce the total alarm burden and potentially reduce the risk of alarm fatigue.
+The downside of applying a machine learning approaching to the aggressive apnea treatment strategy would be that some prolonged apneas may not receive early intervention if the model incorrectly predicted that the apnea will not last for >30 seconds (i.e.
+the false negatives).
+For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier.
+The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds but it does not) to true positives (correctly predting at the 15 second timepoint that the apnea will last for >30 seconds) is high.
+The aim of this study was to determine the accuracy of machine learning models for predicting at the 15 second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
 
@@
 The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds).
-A true positive was an apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds.
-A false positive was an apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds.
 
@@
 
-We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than not intervening for patients who had apnea for >15 seconds (i.e.
-the default strategy).
+We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently used in practice.
+The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the concervative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
 The net benefit is calculated using the formula:
 
-This formula essentially transforms the total number of true positives and false positives into a standardized scale, being the number of true positives, weighted according to the probability threshold.
-For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model.
-This net benefit can result from any combination of true positives and false positives.
+$$
+NB = \frac{TP}{N} - \frac{FP}{N} \times \frac{Pt}{1-Pt}
+$$
 
-A decision curve was created to plot net benefits across a range of probability thresholds, which take into account a plausbile range of clinician preferences for the point at which an intervention should be initiated to treat apnea (this could be, for example, something as simple as verbal stimulation).
-We tested the thresholds in the range from 0.4 to 0.8.
+Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size and $Pt$ is the probability threshold.
+
+This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm or a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model.
+This net benefit can result from any combination of true positives and false positives.[@vickers2016net] The higher the value of $Pt$, the more importance is placed on the number of false positives.
+
+A decision curve was created to plot net benefits across a range of probability thresholds.
+The decision curve was compared against the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) and the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
+The decision curve takes into account a plausbile range of clinician preferences for the point at which an intervention should be initiated to treat apnea (this could be, for example, something as simple as verbal stimulation).
+We tested thresholds in the range from 0.4 to 0.8.
 In a practical sense, this means that we decided that all clinicians would not ever accept a probability for prolonged apnea lower than 0.4 as a useful trigger for intervention, and, conversely, that all clinicians would always consider that intervention is indicated if the probability of prolonged apnea was higher than 0.8.
@@
 
+### Model developmenent and evaluation
+
+Out of sample prediction accuracy for the models using cross-validation is summarized in Table 1. The best model, in terms of
+
 ### Decision curve analysis
 
-Net benefit associated with the random forest model exceeded the strategy of intervening for all patients across all probability thresholds from the range of 0.4 to 0.8.
-The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, would be achieved by using the random forest model as opposed to treating everyone.
-In contrast, the net benefit associated with the random forest model was lower than the strategy of not intervening for any patients across all probability thresholds from the range of 0.4 to 0.8.
-This was due to the fact that we applied a condition that we would not accept more than 5 false positives to detect 1 true positive.
-The number of false positives that triggered across probability thresholds from 0.4 to 0.8 using the random forest model would far exceed the number of true positives, meaning the preferred strategy should remain as only setting the alert for the 30-second time-point.
+Net benefit associated with the random forest model exceeded the strategy of intervening for all patients (the default strategy for the 'aggressive alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2A).
+The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, by using the random forest model.
+The net benefit associated with the random forest model was higher than the strategy of not intervening for any patients at the 15 second mark (the default strategy for the 'conservative alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2B).
 
@@
 Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods.
-It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@Conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
+It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
 Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
@@
 Using the machine learning model would produce a better clinical outcome, in terms of the balance between true positives and false positives for detection of apneic periods over 30 seconds, than a default strategy of intervening for everyone who has an apneic period ≥15 seconds.
+
+## References

### 2021-03-12 _targets.R

@@ -4,11 +4,7 @@ library(tarchetypes)
 devtools::load_all()
 options(tidyverse.quiet = TRUE)
 
-tar_option_set(packages = c(
-  "tidymodels",
-  "tidyverse",
-  "patchwork"
-))
+tar_option_set(packages = c("tidymodels", "tidyverse"))
 list(
   # Load in patient data and respiratory state data
   # and reformat into dataframe
@@ -78,7 +74,7 @@ list(
   tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
 
   # random forest model
-  tar_target(rf, rand_forest(trees = 500) %>%
+  tar_target(rf, rand_forest(trees = 100) %>%
     set_engine("randomForest") %>%
     set_mode("classification")),
   tar_target(
@@ -190,38 +186,7 @@ list(
       `Random forest` = rf_test$.pred_long
     )),
 
-  tar_target(dca_none, dca(
-    data = as.data.frame(dca_data),
-    outcome = "outcome",
-    predictors = "Random forest",
-    smooth = TRUE,
-    xstart = 0.4,
-    xstop = 0.8,
-    graph = F,
-    # wouldn't accept needing to intervene on 5 false postitives to get a true positive
-    harm = 1 / 5
-  )),
-
-  tar_target(dca_plot_none, dca_none$net.benefit %>%
-    select(-`Random forest_sm`,
-      "Threshold" = threshold,
-      -all,
-      "Alarm at ≥30 seconds" = none
-    ) %>%
-    pivot_longer(
-      cols = -Threshold,
-      names_to = "Decisions",
-      values_to = "Net benefit"
-    ) %>%
-    ggplot(aes(x = Threshold, y = `Net benefit`)) +
-    geom_line(aes(linetype = Decisions)) +
-    coord_cartesian(ylim = c(-0.2, 0.2)) +
-    theme_minimal() +
-    theme(
-      legend.position = "bottom",
-      legend.title = element_blank()
-    )),
-  tar_target(dca_all, dca(
+  tar_target(dca, dca(
     data = as.data.frame(dca_data),
     outcome = "outcome",
     predictors = "Random forest",
@@ -231,28 +196,17 @@ list(
     graph = F,
   )),
 
-  tar_target(dca_plot_all, dca_all$net.benefit %>%
+  tar_target(dca_plot, dca$net.benefit %>%
     select(-`Random forest_sm`,
[…]

### 2021-03-12 scratch.R

@@ -60,12 +60,18 @@ coord_cartesian( ylim = c(-0.1, 0.5))
 
 
 rf_test %>%
-    mutate(pred = ifelse(.pred_long > 0.4, "long", "short")) %>%
+    mutate(pred = ifelse(.pred_long > 0.6, "long", "short")) %>%
+        mutate(total_pos = ifelse(pred == "long", 1, 0)) %>%
     mutate(tp = ifelse(pred == "long" & long == "long", 1, 0)) %>%
     mutate(fp = ifelse(pred == "long" & long == "short", 1, 0)) %>%
     summarize(tpos = sum(tp)/nrow(rf_test),
     fpos = sum(fp)/nrow(rf_test),
-    nb = tpos - fpos * (0.4/0.6))
+    nb = tpos - fpos * (0.6/0.4),
+    tp = sum(tp),
+    fp = sum(fp),
+    total_pos = sum(total_pos),
+    afrr_cons = sum(total_pos)/tp,
+    afrr_aggr = 1 - (sum(total_pos)/nrow(rf_test)))
 
 
 

### 2021-03-15 manuscript.Rmd
content

@@
 In practice, two alternative strategies for capnography alarm management are typically used.
-The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (15 seconds).
-The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (30 seconds).
+The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (e.g.
+15 seconds).
+The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (e.g.
+30 seconds).
 
 It is possible that capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
-Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.
+Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]
 
@@
 terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer).
-The prediction algorithm was compared against a default alarm setting for capnography monitors, which triggers an alarm at 30 seconds of apnea.
+The prediction algorithm was compared against typical default alarm settings for capnography monitors.
+These are the aggressive strategy, which triggers an alarm after 15 seconds of apnea, and the conservative strategy, which triggers an alarm after 30 seconds of apnea.
 
@@
 
+Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, dose and type of sedation and analgesia administered) variables were used as predictors.
+Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation.
+Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events.
+A total of 18 variables were used.
+
 ### Statistical analysis
+
+Analyses were performed using R[@rcore2021].
+Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.
+Split into a training and test set containing 75% and 25% of the data, respectively.
 
@@
 #### Modelling
+Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.
+- Normalized predictors
+- interaction terms
 

### 2021-03-16 manuscript.Rmd
multi plots

@@
 For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier.
-The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds but it does not) to true positives (correctly predting at the 15 second timepoint that the apnea will last for >30 seconds) is high.
-The aim of this study was to determine the accuracy of machine learning models for predicting at the 15 second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
+The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second timepoint that the apnea will last for >30 seconds) is high.
+This study aimed to determine the accuracy of machine learning models for predicting at the 15-second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
 
@@
 
-The prediction goal was to classify apneic events at the 15-second time-point as being either short (i.e.
+The prediction goal was to classify apneic events at the 15-second timepoint as being either short (i.e.
 terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer).
@@
 
-Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, dose and type of sedation and analgesia administered) variables were used as predictors.
+Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, body mass index, dose and type of sedation and analgesia administered) variables were used as predictors.
 Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation.
 Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events.
-A total of 18 variables were used.
+A total of 18 predicotr variables were used.
 
@@
 Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.
-Split into a training and test set containing 75% and 25% of the data, respectively.
+
+#### Modelling
+We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression and the XGBoost model.
+Out of sample accuracy of the models was calculated using 10-fold cross-validation.
 
@@
 The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds).
-
-#### Modelling
 Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.
-- Normalized predictors
-- interaction terms
 
@@
 
-We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently used in practice.
+We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently in place.
 The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the concervative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
@@
 
-Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size and $Pt$ is the probability threshold.
+Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size (number of predictions made) and $Pt$ is the probability threshold.
 
-This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm or a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model.
-This net benefit can result from any combination of true positives and false positives.[@vickers2016net] The higher the value of $Pt$, the more importance is placed on the number of false positives.
-
-A decision curve was created to plot net benefits across a range of probability thresholds.
-The decision curve was compared against the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) and the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
-The decision curve takes into account a plausbile range of clinician preferences for the point at which an intervention should be initiated to treat apnea (this could be, for example, something as simple as verbal stimulation).
-We tested thresholds in the range from 0.4 to 0.8.
-In a practical sense, this means that we decided that all clinicians would not ever accept a probability for prolonged apnea lower than 0.4 as a useful trigger for intervention, and, conversely, that all clinicians would always consider that intervention is indicated if the probability of prolonged apnea was higher than 0.8.
-These probability thresholds can be interpreted as value preferences in the context of decision curve analysis.
-For example, a clinician who is 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4).
-Individual participant characteristics will also influence clinicians' decisions about probability thresholds.
-A clinician may elect to intervene when the probability for prolonged apnea is 0.5 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.
-
-A threshold of 0.5 inherently indicates that avoiding a false positive is as valuable of an outcome as identifiying a true positive.
+This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm of a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model.
+This net benefit can result from any combination of true positives and false positives.[@vickers2016net] A probability threshold of 0.5 indicates that avoiding a false positive is as important to a clinician as identifiying a true positive.
 Preferences for probability thresholds below 0.5 are weighted such that identifying a true positive is more valuable than avoiding a false positive.
 Preferences for probability thresholds above 0.5 are weighted such that avoiding a false positive is more valuable than identifying a true positive.
+For example, for a $Pt$ of 0.75, the 'value' of a false positive is three true positives (0.7/0.25 = 3).
+In other words, to create a net benefit from using the model at this probability threshold, there must be more than three true positives for every false positive prediction made from the model.
+Conversely, for a $Pt$ of 0.25, the 'value' of a false positive is weighted far lower, at only a third of a true positive (0.25/0.75 = 0.33).
+This means that a net benefit would be achieved if there was more than one true positive for every three false positives.
 Decision curves can be interpreted such that the strategy with the highest net benefit at each probability threshold has the highest clinical value.[@vickers2016net]
+
+We created a decision curve to plot net benefits across a range of probability thresholds for the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) as well as for the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
+The decision curve takes into account a plausbile range of clinician preferences for the point at which an alarm should be triggered to signal that the patient is apneic.
+We tested thresholds in the range from 0.4 to 0.5 for the agressive alarm management strategy.
+In a practical sense, this means that we decided that all clinicians who usually use the aggressive alarm strategy would not ever accept a probability for prolonged apnea lower than 0.4 as a useful alarm trigger, because there would be little difference between this strategy and just setting the alarm for all apneas.
+We also decided that all clinicians would always consider that an alarm is indicated for the aggresive alarm manamagement strategy if the probability of prolonged apnea was higher than 0.5.
+A range of values were used because these probability thresholds can be interpreted as value preferences in the context of decision curve analysis.
+For example, a clinician who is more 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4).
+Individual participant characteristics will also influence clinicians' decisions about probability thresholds.
+A clinician may elect to intervene when the probability for prolonged apnea is 0.4 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.
+For the conservative alarm management strategy comparison, we chose to plot the range of probability thresholds from 0.7 to 0.8.
+Higher values were chosen because the number of false positives would be an important consideration for clinicians already using a conservative alarm management approach.
 
@@
 
-### Model developmenent and evaluation
+### Discrimination
 
-Out of sample prediction accuracy for the models using cross-validation is summarized in Table 1. The best model, in terms of
+A plot of the area under the receiver operating characteristics curves for the models using predictions from 10-fold cross-validation is presented in Figure 1.
+The random forest  had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03).
+A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value and negative predictive value across all probability thresholds, is presented in Figure 2.
+
+### Calibration
+
+The random forest model had the best callibration.
+It approximated observed risk at moderate (0.5) to high (0.8) thresholds (Figure 3), although risk was overestimated at very low thresholds and slighlty underestimated between 0.4 to 0.5.
+Other models severely overestimated risk at low probability thresholds and underestimated risk at high probability thresholds.
 
@@
 
-Net benefit associated with the random forest model exceeded the strategy of intervening for all patients (the default strategy for the 'aggressive alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2A).
-The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40%, by using the random forest model.
-The net benefit associated with the random forest model was higher than the strategy of not intervening for any patients at the 15 second mark (the default strategy for the 'conservative alarm management approach') across all probability thresholds from the range of 0.4 to 0.8 (Figure 2B).
+As the random forest performed the best in terms of discrimination and callibration, we chose this model to evaluate using decision curve analysis.
+Net benefit associated with the random forest model exceeded the aggressive alarm management strategy across all probability thresholds from the range of 0.4 to 0.5 (Figure 4A).
+The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40% by using the random forest model.
+The net benefit associated with the random forest model was higher than the conservative alarm management strategy across all probability thresholds from the range of 0.7 to 0.8 (Figure 4B).
 
@@
 It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
-Krauss and colleagues[@kraussCharacteristicsPredictorsApnea2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
+Krauss and colleagues[@krauss2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
 It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).
@@
 We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.
+A random forest model had the best discriminitive ability and calibration from those that we tested.
+However, it should be noted that the accuracy of this random forest model was still quite low (AUROC = 0.66).
+Additional research should be undertaken with larger sample sizes to validate our initially promising findings.
 
-Using the machine learning model would produce a better clinical outcome, in terms of the balance between true positives and false positives for detection of apneic periods over 30 seconds, than a default strategy of intervening for everyone who has an apneic period ≥15 seconds.
+It should be noted, though, that many prediction modelling studies focused on clinical outcomes have yielded similarly low AUROC scores.
+For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded a AUROC score of 0.57.[@considine2020vital] When presented with a model with a low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice.
+Based on our results, clinicians using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms.
+This is because using the random forest model would produce a better outcome, in terms of the balance between true positives and false positives, for detection of apneic periods over 30 seconds, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
+
+### Limitations
+
+Cross-validation
+
+Small sample size
+
+Convenience sampling from two sites.
 

### 2021-03-16 R/combine_preds.R

@@ -1,0 +1,24 @@
+#' Combine predictions from models into one dataframe
+#' @importFrom dplyr mutate bind_rows
+#' @export
+
+combine_preds <- function(rf_preds,
+                          boost_preds,
+                          glm_preds,
+                          lasso_preds,
+                          ridge_preds) {
+  rf_preds$model_name <- "Random forest"
+  boost_preds$model_name <- "XGboost"
+  glm_preds$model_name <- "Logistic regression"
+  lasso_preds$model_name <- "Lasso regression"
+  ridge_preds$model_name <- "Ridge regression"
+
+  bind_rows(
+    rf_preds,
+    boost_preds,
+    glm_preds,
+    lasso_preds,
+    ridge_preds
+  ) %>%
+    mutate(outcome = ifelse(long == "long", 1, 0))
+}

### 2021-03-16 _targets.R

@@ -71,10 +71,10 @@ list(
   tar_target(wf, workflow() %>%
     add_recipe(base_rec)),
 
-  tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
+  # tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
 
   # random forest model
-  tar_target(rf, rand_forest(trees = 100) %>%
+  tar_target(rf, rand_forest(trees = 500) %>%
     set_engine("randomForest") %>%
     set_mode("classification")),
   tar_target(
@@ -90,53 +90,52 @@ list(
   tar_target(rf_preds, rf_fit %>%
     collect_predictions()),
 
-  tar_target(rf_ppv, rf_preds %>%
-    mutate(pred = ifelse(.pred_long > 0.7, "long", "short")) %>%
-    ppv(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)),
-
-  tar_target(rf_sens, rf_preds %>%
-    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
-    sensitivity(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)),
-
-  tar_target(rf_f_meas, rf_preds %>%
-    mutate(pred = ifelse(.pred_long > 0.3, "long", "short")) %>%
-    f_meas(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)),
-  tar_target(rf_precision, rf_preds %>%
-    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
-    precision(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)),
-  tar_target(rf_recall, rf_preds %>%
-    mutate(pred = ifelse(.pred_long > 0.5, "long", "short")) %>%
-    recall(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)),
-
 
 
   # Only when optimized feature selection and tuning
-  tar_target(rf_test, last_fit(
-    wf %>%
-      add_model(rf),
-    init_split
-  ) %>%
-    collect_predictions()),
+  # if using train/test split
+  # tar_target(rf_test, last_fit(
+  #   wf %>%
+  #     add_model(rf),
+  #   init_split
+  # ) %>%
+  #   collect_predictions()),
+
+  tar_target(combined_preds, combine_preds(
+    rf_preds,
+    boost_preds,
+    glm_preds,
+    lasso_preds,
[…]

### 2021-03-16 scratch.R

@@ -59,23 +59,30 @@ coord_cartesian( ylim = c(-0.1, 0.5))
 
 
 
-rf_test %>%
+rf_preds %>%
     mutate(pred = ifelse(.pred_long > 0.6, "long", "short")) %>%
         mutate(total_pos = ifelse(pred == "long", 1, 0)) %>%
+        mutate(pos = ifelse(long == "long", 1, 0)) %>%
     mutate(tp = ifelse(pred == "long" & long == "long", 1, 0)) %>%
     mutate(fp = ifelse(pred == "long" & long == "short", 1, 0)) %>%
-    summarize(tpos = sum(tp)/nrow(rf_test),
+    summarize(
+        max = max(.pred_long),
+        tpos = sum(tp)/nrow(rf_test),
     fpos = sum(fp)/nrow(rf_test),
     nb = tpos - fpos * (0.6/0.4),
     tp = sum(tp),
     fp = sum(fp),
     total_pos = sum(total_pos),
+   pos = sum(pos),
     afrr_cons = sum(total_pos)/tp,
     afrr_aggr = 1 - (sum(total_pos)/nrow(rf_test)))
 
 
 
-    ppv(estimate = factor(pred), truth = long) %>%
-    pull(.estimate)
+threshperf_plot_multi(test %>%
+mutate(outcome = ifelse(long == "long", 1, 0)),
+outcome = "outcome", 
+prediction = ".pred_long",
+model = "model_name")
 
 

### 2021-03-18 manuscript.Rmd
need to add participant table

@@
 ---
+
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE)
+```
 
@@
 
+A total of 384 apneic events of at least 15 seconds duration from 61 patients who had a procedure with sedation were included in this analysis.
+A summary of participant characteristics is presented in Table 1.
+
 ### Discrimination
@@
 A plot of the area under the receiver operating characteristics curves for the models using predictions from 10-fold cross-validation is presented in Figure 1.
+
+```{r}
+targets::tar_read(roc_plot)
+```
+
 The random forest  had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03).
 A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value and negative predictive value across all probability thresholds, is presented in Figure 2.
+
+```{r}
+targets::tar_read(tpp)
+```
 
@@
 Other models severely overestimated risk at low probability thresholds and underestimated risk at high probability thresholds.
+
+```{r}
+targets::tar_read(cal_plot)
+```
 
@@
 
-It should be noted, though, that many prediction modelling studies focused on clinical outcomes have yielded similarly low AUROC scores.
+Many prediction modelling studies focused on predicting clinical outcomes have yielded similarly low AUROC scores.
 For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded a AUROC score of 0.57.[@considine2020vital] When presented with a model with a low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice.
-Based on our results, clinicians using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms.
-This is because using the random forest model would produce a better outcome, in terms of the balance between true positives and false positives, for detection of apneic periods over 30 seconds, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
+Based on our results, nurses currently using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+This is because using the random forest model would produce a better outcome, in terms of the balance between true positives and false positives, for detecting if an apneic event will be prolonged, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
+Likewise, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+This strategy would reduce the total alarm burden.
 
@@
 
-Cross-validation
+Although the number of apneas included in the models was high, the overall sample size for this study was small (n=102). Cross-validation
 

### 2021-03-18 manuscript.Rmd
basic table added

@@
 Out of sample accuracy of the models was calculated using 10-fold cross-validation.
+Many participants in the study contributed multiple apneic events to the dataset used for modelling.
+To take this dependency into account, we ensured that apneic events from individual participants were not ever included in both the training and testing partitions of the 10-fold cross-validation process.
 
@@
 
-A total of 384 apneic events of at least 15 seconds duration from 61 patients who had a procedure with sedation were included in this analysis.
+A total of 384 apneic events of at least 15 seconds duration from 61 of the 102 patients who participated in this observational study were included in the present analysis.
 A summary of participant characteristics is presented in Table 1.
+
+```{r}
+targets::tar_read(summary_table)
+```
 
@@
 
-Although the number of apneas included in the models was high, the overall sample size for this study was small (n=102). Cross-validation
+Although the number of apneas included in the models was relative high, these were contributed by a small number of patients (n=61).
+We used cross-validation to minimize the possibility of overfitting.
+It should also be noted that the data used in this analysis were produced from an observation study conducted at jsut two hospitals that used a convenience sampling approach.
+Therefore the possibility of selection bias should be considered.
+The context in which the study was conducted should also be considered in regard to external validity.
+Participants in the study were patients undergoing procedures in a cardiac catheterization laboratory where small, bolus doses of midazolam and fentanyl were used for sedation.
+Other procedural sedation contexts may use different doses of sedation and the types of medications.
+It would not be expected that the results of this study could be generalized to those other contexts.
 
-Small sample size
-
-Convenience sampling from two sites.
+### Conclusion
 

### 2021-03-18 R/summary_table.R

@@ -1,0 +1,17 @@
+#' Create a summary table of participant characteristics
+#' @importFrom gtsummary tbl_summary
+#' @importFrom dplyr select group_by filter
+#' @param combined dataframe used for modelling
+#' @export 
+#' 
+create_summary_table <- function(combined) {
+    combined %>%
+    group_by(ID) %>%
+    filter(row_number() == 1) %>%
+    ungroup() %>%
+    select(Age, "Body mass index" = BMI,
+    Sex,
+    "Obstructive sleep apnea" = OSA,
+    "American Society of Anesthesiology Physical Classification Status" = ASA) %>%
+    tbl_summary()
+}

### 2021-03-18 _targets.R

@@ -23,6 +23,11 @@ list(
     upper_trigger = 30
   )),
 
+  tar_target(
+    summary_table,
+    create_summary_table(combined)
+  ),
+
   tar_target(data_split, combined %>%
     make_train_id(
       seed = 42,
@@ -112,7 +117,7 @@ list(
   tar_target(
     tpp,
     runway::threshperf_plot_multi(
-     combined_preds,
+      combined_preds,
       outcome = "outcome",
       prediction = ".pred_long",
       model = "model_name"

### 2021-03-19 manuscript.Rmd
full first draft

--- 2021-03-16 manuscript.Rmd
@@
 ---
-title: "Predicting prolonged apneic episodes using machine learning during nurse-administered procedural sedation"
+title: "Predicting prolonged apnea during nurse-administered procedural sedation using machine learning"
 bibliography: "Extras/references.bib"
+csl: "Extras/anaesthesia.csl"
+# output: distill::distill_article
+author:
+  - Aaron Conway PhD:
+      email: aaron.conway@utoronto.ca
+      phone: +1 416 946 7112
+      address: 585 University Ave | Toronto, ON  M5G 2N2
+      institute: [PMCC, LBFON, QUT]
+      correspondence: true
+  - Carla Jungquist PhD:
+      institute: [UoB]
+      correspondence: false
+  - Kristina Chang MScN:
+      institute: [PMCC]
+      correspondence: false
+  - Navpreet Kamboj BScN:
+      institute: [LBFON]
+      correspondence: false
+  - Joanna Sutherland MBBS:
+      institute: [NSW, Coffs]
+      correspondence: false
+  - Sebastian Mafeld:
+      institute: [IR]
+      correspondence: false
+  - Matteo Parotto MD, PhD:
+      institute: [TGH, UoT]
+      correspondence: false
+institute:
+  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
+  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
+  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
+  - UoB: The University of Buffalo, Buffalo, New York, USA
+  - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
+  - Coffs: Department of Anaesthesia, Coffs Harbour Health Campus, Coffs Harbour, NSW, Australia
+  - IR: Interventional Radiology, JDMI, Toronto General Hospital
+  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
+  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
+output:
+  officedown::rdocx_document:
+    reference_docx: Extras/manuscript_template.docx
+    pandoc_args:
+      - --lua-filter=Extras/scholarly-metadata.lua
+      - --lua-filter=Extras/author-info-blocks.lua
 ---
+
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE)
+```
+
+\pagebreak
+
+# Abstract
+
+## Introduction
+
+Capnography is commonly used for nurse-administered procedural sedation.
+This study aimed to determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will persist for 30 seconds or more.
+This information would be useful for determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
+
+## Methods
+
+A secondary analysis of an observational study was conducted.
+We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression and the XGBoost model.
+Out of sample accuracy of the models was calculated using 10-fold cross-validation.
+The net benefit decision analytic measure was used to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently in place.
+The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the conservative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
+
+## Results
+
+A total of 384 apneas longer than 15 seconds were observed in 61 of the 102 patients who participated in the observational study.
+Nearly half of the apneas (n=180) were prolonged.
+The random forest had the best discrimination (AUROC 0.66) and calibration.
+Net benefit associated with the random forest model exceeded the aggressive alarm management strategy but was lower than the conservative alarm management strategy.
+
+## Conclusion
+
+Decision curve analysis indicated that using a random forest model would lead to a better outcome for capnography alarm management in comparison to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
+The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
+
+\pagebreak
 
@@
 Capnography waveform abnormalities assist in the detection and diagnosis of specific conditions, such as partial airway obstruction and apnea.
-For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities including peak professional organisations for anaesthesia in Canada, the United States and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018]
+For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities, including peak professional organizations for anesthesia in Canada, the United States, and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018] It is commonly used for nurse-administered procedural sedation, [@Conway_2014; @conway2013; @conway2013a] including in the interventional radiology setting.[@pella2018systematic; @brast2016capnography; @long2016capnography; @mafeld2020avoiding; @werthman2020administration]
 
@@
 Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice.
-For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases risk of alarm fatigue.
+For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue.
 Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation.
@@
 For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier.
-The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second timepoint that the apnea will last for >30 seconds) is high.
+The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second time point that the apnea will last for >30 seconds) is high.
 This study aimed to determine the accuracy of machine learning models for predicting at the 15-second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
 
-# Methods
+## Methods
 
@@
 The capnography waveform was displayed on the main physiological monitoring screen.
-A default ‘No breaths detected’ alert was triggered for apnea but no other audible or visual alarms were set for the capnography monitor.
-No restrictions or specific instructions were provided to clinicians regarding the detection of capnography waveform abnormalities as part of the research protocol because the study used observational design.
+A default ‘No breaths detected’ alert was triggered for apnea, but no other audible or visual alarms were set for the capnography monitor.
+No restrictions or specific instructions were provided to clinicians regarding the detection of capnography waveform abnormalities as part of the research protocol because the study used an observational design.
 
@@
 Several raw demographic (age, sex) and clinical (ASA score, diagnosis of sleep apnea, body mass index, dose and type of sedation and analgesia administered) variables were used as predictors.
-Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and time since the previous dose of sedation.
-Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event and total number of apneic events.
-A total of 18 predicotr variables were used.
+Features related to sedation dosing used as predictors in the model were the total dose of sedation and number of doses of sedation administered, time since first sedation and the time since the previous dose of sedation.
+Other features were extracted from the capnography waveform to use as predictors, such as the previous respiratory state (defined as either normal breathing or abnormal), duration of the previous apneic event, time since the previous apneic event, and the total number of apneic events.
+A total of 18 predictor variables were used.
 
@@
 #### Modelling
-We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression and the XGBoost model.
+We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression, and the XGBoost model.
 Out of sample accuracy of the models was calculated using 10-fold cross-validation.
-
-The best model was chosen with respect to the optimal balance between maximising the ability to predict prolonged events and alarm frequency.
-The alarm frequency metric used was the ratio of correct predictions of short apneic periods to the total number of apneic periods that triggered alerts, which was calculated as TP/(TP + FP + apneas > 30seconds).
-Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and total number of apneic events.
+Many participants in the study contributed multiple apneic events to the dataset used for modeling.
+To take this dependency into account, we ensured that apneic events from individual participants were not included in both the training and testing partitions of the 10-fold cross-validation process.
+Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and the total number of apneic events.
 
@@
 We used the net benefit decision analytic measure to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently in place.
-The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the concervative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
+The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the conservative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
 The net benefit is calculated using the formula:
@@
 
-Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total of number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size (number of predictions made) and $Pt$ is the probability threshold.
+Where $NB$ is the net benefit, $TP$ is the total number of true positives (apneic event at 15 seconds which was predicted to be prolonged and did persist for >30 seconds), $FP$ is the total number of false positives (apneic events at 15 seconds that was predicted to be prolonged but did not persist for >30 seconds), $N$ is the sample size (number of predictions made), and $Pt$ is the probability threshold.
 
-This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm of a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be 7 true positives from every 100 predictions from the model.
-This net benefit can result from any combination of true positives and false positives.[@vickers2016net] A probability threshold of 0.5 indicates that avoiding a false positive is as important to a clinician as identifiying a true positive.
+This formula essentially transforms the total number of true positives and false positives into a standardized scale, weighted by the relative harm of a false-positive result.[@vickers2006decision] For example, a net benefit of 0.07 means the net benefit of using the model would be seven true positives from every 100 predictions from the model.
+This net benefit can result from any combination of true positives and false positives.[@vickers2016net] A probability threshold of 0.5 indicates that avoiding a false positive is as important to a clinician as identifying a true positive.
 Preferences for probability thresholds below 0.5 are weighted such that identifying a true positive is more valuable than avoiding a false positive.
@@
 We created a decision curve to plot net benefits across a range of probability thresholds for the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) as well as for the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
-The decision curve takes into account a plausbile range of clinician preferences for the point at which an alarm should be triggered to signal that the patient is apneic.
-We tested thresholds in the range from 0.4 to 0.5 for the agressive alarm management strategy.
-In a practical sense, this means that we decided that all clinicians who usually use the aggressive alarm strategy would not ever accept a probability for prolonged apnea lower than 0.4 as a useful alarm trigger, because there would be little difference between this strategy and just setting the alarm for all apneas.
-We also decided that all clinicians would always consider that an alarm is indicated for the aggresive alarm manamagement strategy if the probability of prolonged apnea was higher than 0.5.
-A range of values were used because these probability thresholds can be interpreted as value preferences in the context of decision curve analysis.
-For example, a clinician who is more 'risk-averse' may elect for a more conservative probability threshold (closer to 0.4).
+The decision curve takes into account the full range of reasonable clinician preferences for the point at which an alarm should be triggered to signal that the patient is apneic.
+We tested thresholds in the range from 0.3 to 0.5 for the aggressive alarm management strategy.
+In a practical sense, this means that we decided that all clinicians who usually use the aggressive alarm strategy would not ever accept a probability for prolonged apnea lower than 0.3 as a useful alarm trigger because there would be little difference between this strategy and just setting the alarm for all apneas.
+We also decided that all clinicians would always consider that an alarm is indicated for the aggressive alarm management strategy if the probability of prolonged apnea was higher than 0.5.
+A range of values was used because these probability thresholds can be interpreted as value preferences that individual clinicians may reasonably choose in the context of clinical practice.
+For example, a clinician who is more 'risk-averse' may elect for a more conservative probability threshold (closer to 0.3).
 Individual participant characteristics will also influence clinicians' decisions about probability thresholds.
-A clinician may elect to intervene when the probability for prolonged apnea is 0.4 for an elderley patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.
+A clinician may elect to intervene when the probability for prolonged apnea is 0.3 for an elderly patient with multiple comorbidities but not for a young patient who may be more likely to be able to tolerate longer periods of apnea.
 For the conservative alarm management strategy comparison, we chose to plot the range of probability thresholds from 0.7 to 0.8.
@@
 
+A total of 384 apneic events of at least 15 seconds duration from 61 of the 102 patients who participated in this observational study were included in the present analysis.
+A summary of participant characteristics is presented in Table \@ref(tab:tab1).
+Nearly half of the (n=180) of the apneic events were prolonged, indicating a balanced dataset for the prediction problem at hand.
+
 ### Discrimination
 
-A plot of the area under the receiver operating characteristics curves for the models using predictions from 10-fold cross-validation is presented in Figure 1.
+A plot of the area under the receiver operating characteristics curves for the models using predictions from 10-fold cross-validation is presented in Figure \@ref(fig:roc).
+
 The random forest had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03).
-A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value and negative predictive value across all probability thresholds, is presented in Figure 2.
+A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value, and negative predictive value across all probability thresholds, is presented in Figure \@ref(fig:tpp).
 
@@
 
-The random forest model had the best callibration.
-It approximated observed risk at moderate (0.5) to high (0.8) thresholds (Figure 3), although risk was overestimated at very low thresholds and slighlty underestimated between 0.4 to 0.5.
+The random forest model had the best calibration.
+It approximated observed risk at moderate (0.5) to high (0.8) thresholds (Figure \@ref(fig:cal)), although the risk was overestimated at very low thresholds and slightly underestimated between 0.4 to 0.5.
 Other models severely overestimated risk at low probability thresholds and underestimated risk at high probability thresholds.
@@
 
-As the random forest performed the best in terms of discrimination and callibration, we chose this model to evaluate using decision curve analysis.
-Net benefit associated with the random forest model exceeded the aggressive alarm management strategy across all probability thresholds from the range of 0.4 to 0.5 (Figure 4A).
-The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15 second mark if the probability of it being prolonged was more than 40% by using the random forest model.
-The net benefit associated with the random forest model was higher than the conservative alarm management strategy across all probability thresholds from the range of 0.7 to 0.8 (Figure 4B).
+As the random forest performed the best in terms of discrimination and calibration, we chose this model to evaluate using decision curve analysis.
+The net benefit associated with the random forest model exceeded the aggressive alarm management strategy across all probability thresholds from the range of 0.3 to 0.5 (Figure \@ref(fig:dca)).
+The interpretation is that the best clinical outcome would be achieved for clinicians who are willing to initiate an intervention to treat apnea at the 15-second mark if the probability of it being prolonged was more than 40% by using the random forest model.
+The net benefit associated with the random forest model was lower than the conservative alarm management strategy across all probability thresholds from the range of 0.7 to 0.8 (Figure \@ref(fig:dca)).
 
@@
 We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.
-A random forest model had the best discriminitive ability and calibration from those that we tested.
+A random forest model had the best discriminative ability and calibration from those that we tested.
 However, it should be noted that the accuracy of this random forest model was still quite low (AUROC = 0.66).
@@
 
-It should be noted, though, that many prediction modelling studies focused on clinical outcomes have yielded similarly low AUROC scores.
-For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded a AUROC score of 0.57.[@considine2020vital] When presented with a model with a low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice.
-Based on our results, clinicians using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms.
-This is because using the random forest model would produce a better outcome, in terms of the balance between true positives and false positives, for detection of apneic periods over 30 seconds, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
+Many prediction modeling studies focused on predicting clinical outcomes have yielded similarly low AUROC scores.
+For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded an AUROC score of 0.57.[@considine2020vital] When presented with a model with low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice.
+Based on our results, nurses currently using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would not derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+This is because using the random forest model would produce a worse outcome, in terms of the balance between true positives and false positives, for detecting if an apneic event will be prolonged, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
+
+Conversely, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+Using the random forest model as an additional input for an alarm trigger would reduce the total alarm burden and could be considered as an option for implementation into practice.
+To operationalize these predictions into capnography monitors, partnerships with industry would be required.[@glasgow2018nurse] Integrating predictive models into alarm management strategies for respiratory monitoring devices is also indicated in other contexts.
+For example, a recent study found that opioid-induced respiratory depression during recovery from anesthesia can be accurately predicted using a machine learning approach.[@jungquist2019identifying] In addition, user-centered design considerations, such as how the predictions should be communicated to nurses responsible for decision-making, would be important avenues for further research prior to implementation.[@risling2020advancing]
+
+The finding that the model was able to produce an overall net benefit against the aggressive capnography alarm management strategy but not the conservative approach is a noteworthy finding.
+Further research with larger sample sizes is needed to increase the predictive power of models aimed at predicting the duration of apnea.
+Such research is warranted because triggering an alarm after 30 seconds of apnea that would have self-resolved without clinical intervention only 5 seconds later is just as clinically inconsequential as triggering an alert after 5 seconds of apnea that would similarly have resolved after a short period of time.
+In both of these circumstances, there would not have been enough time for the clinician's intervention to take effect.
+Yet, presumably in an attempt to reduce alarm burden, many capnography monitoring devices alarm settings default to triggering after apnea has been observed for 30 seconds.
+An ideal alternative to the conservative alarm strategy would be for clinicians to receive capnography monitor alarms as _early as possible_ in the course of an apneic event, _but only_ if the event will be prolonged for enough time for a clinical intervention to be implemented and exert its intended effect.
+Results from the present study did not achieve this goal.
+Yet, previous research indicates that finding such a solution may be worthwhile.
+An analysis of over 500 thousand patients found that respiratory compromise in interventional radiology procedures performed with moderate sedation contributed to worse clinical outcomes and higher costs.[@urman2019impact]
 
@@
 
-Cross-validation
+Although the number of apneas included in the models was relatively high, these were contributed by a small number of patients.
+We used cross-validation to minimize the possibility of overfitting.
+It should also be noted that the data used in this analysis were produced from an observational study conducted at two hospitals that used a convenience sampling approach.
+Therefore the possibility of selection bias should be considered.
+The context in which the study was conducted should also be considered in regard to external validity.
+Participants in the study were patients undergoing procedures in a cardiac catheterization laboratory where small, bolus doses of midazolam and fentanyl were used for sedation.
+Other procedural sedation contexts may use different doses of sedation and the types of medications.
+It would not be expected that the results of this study could be generalized to those other contexts.
 
-Small sample size
+### Conclusion
 
-Convenience sampling from two sites.
+We evaluated several candidate models to determine their accuracy in predicting if an apnea at 15 seconds would be prolonged for more than 30 seconds.
+The random forest model provided the best discrimination and calibration.
+Decision curve analysis indicated that using the random forest model would lead to a better outcome for capnography alarm management in comparison to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
+The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
+
+\pagebreak
 
 ## References
+
+::: {#refs}
+:::
+
+\pagebreak
+
+## Figure legend
+
+Fig. 1 Area under receiver operating characteristics curve
+
+Fig. 2 Threshold performance plot for all models evaluated
+
+Fig. 3 Calibration plot for all models evaluated
+
+Fig. 4 Decision curve analysis plots. Panel A is the comparison for the aggressive alarm management strategy and panel B is the comparison for the conservative alarm management strategy.
+
+\pagebreak
+
+```{r tab.cap='Participant characteristics', tab.id = 'tab1'}
+targets::tar_read(summary_table)
+```
+
+\pagebreak
+
+```{r roc, fig.cap="Area under receiver operating characteristics curve"}
+targets::tar_read(roc_plot)
+```
+
+<!---BLOCK_MULTICOL_START--->
+
+<!---BLOCK_STOP--->
+
+<!--BLOCK_LANDSCAPE_START--->
+
+```{r tpp, fig.cap="Threshold performance plot for all models evaluated", fig.width=9}
+targets::tar_read(tpp)
+```
+
+\pagebreak
+
+```{r cal, fig.cap="Calibration plot for all models evaluated"}
+targets::tar_read(cal_plot)
+```
+\pagebreak
+
+```{r dca, fig.cap="Decision curve analysis plots.
+Panel A is the comparison for the aggressive alarm management strategy and panel B is the comparison for the conservative alarm management strategy", fig.width=7}
+targets::tar_read(dca_plot_combined)
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->

### 2021-03-19 R/format_data.R

@@ -83,8 +83,8 @@ format_data <- function(resp_data, durations,
         )) %>%
         na.omit() %>%
         mutate(ASA = case_when(
-            ASA == 1 | ASA == 2 ~ "Low",
-            TRUE ~ "High"
+            ASA == 1 | ASA == 2 ~ "I or II",
+            TRUE ~ "III or IV"
         )) %>%
         mutate(last_state = case_when(
             last_state == "Normal breathing" ~ "Normal breathing",

### 2021-03-19 _targets.R

@@ -4,7 +4,10 @@ library(tarchetypes)
 devtools::load_all()
 options(tidyverse.quiet = TRUE)
 
-tar_option_set(packages = c("tidymodels", "tidyverse"))
+tar_option_set(packages = c(
+  "tidymodels", "tidyverse",
+  "patchwork"
+))
 list(
   # Load in patient data and respiratory state data
   # and reformat into dataframe
@@ -79,7 +82,7 @@ list(
   # tar_target(class_metrics, metric_set(ppv, sensitivity, specificity)),
 
   # random forest model
-  tar_target(rf, rand_forest(trees = 500) %>%
+  tar_target(rf, rand_forest(trees = 100) %>%
     set_engine("randomForest") %>%
     set_mode("classification")),
   tar_target(
@@ -237,20 +240,22 @@ list(
     select(-`Random forest_sm`,
       "Threshold" = threshold,
       -all,
-      "Alarm at 30 seconds of apnea" = none
+      "Alarm at 30 seconds" = none
     ) %>%
     pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
     ggplot(aes(x = Threshold, y = `Net benefit`)) +
-    geom_line(aes(linetype = Decisions)) +
+    geom_line(aes(color = Decisions)) +
     # coord_cartesian(ylim = c(-0.05, 0.1)) +
     theme_minimal() +
-    theme(legend.position = "bottom")),
+    theme(legend.position = "bottom",
+      legend.title = element_blank())),
+
   tar_target(dca15, dca(
     data = as.data.frame(dca_data),
     outcome = "outcome",
     predictors = "Random forest",
     smooth = TRUE,
-    xstart = 0.4,
+    xstart = 0.3,
     xstop = 0.5,
     graph = F,
   )),
@@ -258,15 +263,24 @@ list(
   tar_target(dca_plot15, dca15$net.benefit %>%
     select(-`Random forest_sm`,
       "Threshold" = threshold,
-      "Alarm at 15 seconds of apnea" = all,
+      "Alarm at 15 seconds" = all,
       -none
     ) %>%
     pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
     ggplot(aes(x = Threshold, y = `Net benefit`)) +
-    geom_line(aes(linetype = Decisions)) +
+    geom_line(aes(color = Decisions)) +
[…]

### 2021-03-19 scratch.R

@@ -67,15 +67,16 @@ rf_preds %>%
     mutate(fp = ifelse(pred == "long" & long == "short", 1, 0)) %>%
     summarize(
         max = max(.pred_long),
-        tpos = sum(tp)/nrow(rf_test),
-    fpos = sum(fp)/nrow(rf_test),
+        tpos = sum(tp)/nrow(rf_preds),
+    fpos = sum(fp)/nrow(rf_preds),
     nb = tpos - fpos * (0.6/0.4),
     tp = sum(tp),
     fp = sum(fp),
     total_pos = sum(total_pos),
+   pos_pr = sum(pos)/nrow(rf_preds),
    pos = sum(pos),
     afrr_cons = sum(total_pos)/tp,
-    afrr_aggr = 1 - (sum(total_pos)/nrow(rf_test)))
+    afrr_aggr = 1 - (sum(total_pos)/nrow(rf_preds)))
 
 
 
@@ -86,3 +87,6 @@ prediction = ".pred_long",
 model = "model_name")
 
 
+data_split %>%
+distinct(ID) %>%
+count()

### 2021-03-24 Extras/cover-letter.Rmd
co-author updates and jrn docs

---
date: "`r Sys.Date()`"
toaddress: Journal of Radiology Nursing
toname: Editorial Committee
output: letters::uhn_letter
knit: pagedown::chrome_print
---

On behalf of my co-authors, I am pleased to submit our manuscript entitled “Predicting prolonged apnea during nurse-administered procedural sedation using machine learning".

Capnography is commonly used for nurse-administered procedural sedation, including in interventional radiology. Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementating this technology into practice. For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue. Conversely, only intervening once an apneic period reaches a longer threshold negates capnography's potential benefits on patient safety by improving ventilation. In practice, two alternative strategies for capnography alarm management are typically used. The 'aggressive' strategy is for alarms to be triggered after short periods of apnea (e.g. 15 seconds). The 'conservative' approach triggers the alarm only once the patient has been apneic for a prolonged period (e.g. 30 seconds). This study aimed to determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will persist for 30 seconds or more. This information would help determine whether operationalizing these predictions in practice as alarm triggers would be beneficial.

We conducted a secondary analysis of a prospective observational study. The primary aim of the observational study was to identify common patterns in capnography waveform abnormalities and factors that influence these patterns. Results for the primary aim of the observational study are reported elsewhere. We believe this secondary analysis makes a useful additional contribution to the evidence base, as it indicates the potential utility of a machine learning approach for optimizing alarm management during procedural sedation. 
This approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.

### 2021-03-24 manuscript.Rmd
co-author updates and jrn docs

@@
 csl: "Extras/anaesthesia.csl"
-# output: distill::distill_article
 author:
@@
   - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
-  - UoT: Department of Anesthesia and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
+  - UoT: Department of Anesthesiology and Pain Medicine and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
 output:
@@
 
+\pagebreak
+
+Declarations of interest: none
+
+Funding: A.C.
+received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S.
+Bloomberg Faculty of Nursing at the University of Toronto.
+This study received funding from the Wesley Medical Research (Project number: 2015-31).
+
+Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.
+
 ```{r setup, echo=FALSE}
-knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(echo = FALSE,message=FALSE, warning=FALSE, fig.retina=8)
 ```
@@
 
-# Abstract
+## Abstract
 
@@
 This study aimed to determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will persist for 30 seconds or more.
-This information would be useful for determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
+This information would help determine whether operationalizing these predictions in practice as alarm triggers would be beneficial.
 
@@
 
-Decision curve analysis indicated that using a random forest model would lead to a better outcome for capnography alarm management in comparison to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
+Decision curve analysis indicated that using a random forest model would lead to a better outcome for capnography alarm management compared to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
 The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
@@
 
+With the recent increase in electronic monitoring devices in the hospital setting, alarm fatigue has become a serious problem that impacts patient safety and nursing care.[@lewandowska2020impact] Alarm fatigue is caused by exposure to excessive and frequent device alarms and consequently becoming desensitized to them.
+Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] One of the sources of alarms is the capnography device that is used to measure and monitor patients’ ventilation.
+
 A capnography waveform displays the level of expired carbon dioxide (CO~2~) over time to show changes in concentrations throughout the respiratory cycle.
-Capnography waveform abnormalities assist in the detection and diagnosis of specific conditions, such as partial airway obstruction and apnea.
-For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities, including peak professional organizations for anesthesia in Canada, the United States, and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018] It is commonly used for nurse-administered procedural sedation, [@Conway_2014; @conway2013; @conway2013a] including in the interventional radiology setting.[@pella2018systematic; @brast2016capnography; @long2016capnography; @mafeld2020avoiding; @werthman2020administration]
+Capnography waveform abnormalities assist in detecting and diagnosing specific conditions, such as partial airway obstruction and apnea.
+For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities, including national and international professional organizations for anesthesiology in Canada, the United States, and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018] It is commonly used for nurse-administered procedural sedation, [@Conway_2014; @conway2013; @conway2013a] including in the interventional radiology setting.[@pella2018systematic; @brast2016capnography; @long2016capnography; @mafeld2020avoiding; @werthman2020administration]
 
-A potential consequence to consider in regard to the implementation of capnography for patient monitoring in inpatient wards is alarm fatigue.
-Alarm fatigue has been linked to patient deaths resulting from clinicians becoming desensitized to alarms leading to delayed responses to clinical deterioration.[@chopra2014redesigning] Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successful implementation of this technology into practice.
+Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementating this technology into practice.
 For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue.
-Conversely, only intervening once an apneic period reaches a longer threshold negates the potential benefits of capnography on patient safety by improving ventilation.
+Conversely, only intervening once an apneic period reaches a longer threshold negates capnography's potential benefits on patient safety by improving ventilation.
 In practice, two alternative strategies for capnography alarm management are typically used.
@@
 
-It is possible that capnography monitor alarm management may be improved upon by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
+It is possible that capnography monitor alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
 Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]
@@
 For the aggressive strategy, only triggering an alarm for apneas predicted to be prolonged would reduce the total alarm burden and potentially reduce the risk of alarm fatigue.
-The downside of applying a machine learning approaching to the aggressive apnea treatment strategy would be that some prolonged apneas may not receive early intervention if the model incorrectly predicted that the apnea will not last for >30 seconds (i.e.
+The downside of applying a machine learning approach to the aggressive apnea treatment strategy would be that some prolonged apneas may not receive early intervention if the model incorrectly predicted that the apnea will not last for >30 seconds (i.e.
 the false negatives).
 For the conservative strategy, triggering an alarm at the 15-second timepoint for apneas predicted to be prolonged could reduce the total time of apnea, as treatment could be initiated earlier.
-The downside of applying a machine learning approaching to the conservative apnea treatment strategy would be that there is the potential that it could increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second time point that the apnea will last for >30 seconds) is high.
-This study aimed to determine the accuracy of machine learning models for predicting at the 15-second timepoint if a period of apnea will persist for 30 seconds or more with the goal of determining whether operationalizing these predictions in practice as alarm triggers would be beneficial.
+The downside of applying a machine learning approach to the conservative apnea treatment strategy would be the potential to increase the total alarm burden if the ratio of false positives (predicting the apnea will last for >30 seconds, but it does not) to true positives (correctly predicting at the 15-second time point that the apnea will last for >30 seconds) is high.
+This study aimed to determine the accuracy of machine learning models for predicting at the 15-second timepoint if a period of apnea will persist for 30 seconds or more.
+This information would help to determine whether operationalizing these predictions in practice as alarm triggers would be beneficial.
 
@@
 A secondary analysis of a prospective observational study was undertaken.
+The primary aim of the observational study was to identify common patterns in capnography waveform abnormalities and factors that influence these patterns.
 Results for the primary aim of the observational study are reported elsewhere.[@conway2019sequence] All participants provided written informed consent and the study was approved by human research ethics committees (UCH HREC 1614; SVHAC HREC 16/26; QUT 1600000641).
@@
 
-The prediction goal was to classify apneic events at the 15-second timepoint as being either short (i.e.
+The prediction goal was to classify apneic events at the 15-second timepoint as either short (i.e.
 terminated prior to 30 seconds) or prolonged (persisted for 30 seconds or longer).
 The prediction algorithm was compared against typical default alarm settings for capnography monitors.
-These are the aggressive strategy, which triggers an alarm after 15 seconds of apnea, and the conservative strategy, which triggers an alarm after 30 seconds of apnea.
 
@@
 Participants in the study were consecutive adult patients scheduled to undergo an elective procedure in the cardiac catheterization laboratory with moderate sedation.
-Patients with severe cognitive impairment (due to inability to provide informed consent) or inability to understand and speak English (if an interpreter was unavailable) were excluded.
+Patients with severe cognitive impairment (due to inability to provide informed consent) or unable to understand and speak English (if an interpreter was unavailable) were excluded.
+Data collection was performed at two urban private hospitals in Australia.
 
@@
 Sedation was administered by nurses who were trained in advanced life support.
-Routine clinical monitoring included cardiac rhythm and pulse oximetry monitoring as well as non-invasive blood pressure measurements every 5-10 minutes.
+Routine clinical monitoring included continuous cardiac rhythm and oxygen saturation monitoring as well as non-invasive blood pressure measurements every 5-10 minutes.
 The Respironics LoFlo sidestream CO~2~ sensor was used for capnography monitoring.
@@
 Preferences for probability thresholds above 0.5 are weighted such that avoiding a false positive is more valuable than identifying a true positive.
-For example, for a $Pt$ of 0.75, the 'value' of a false positive is three true positives (0.7/0.25 = 3).
+For example, for a $Pt$ of 0.75, the 'value' of a false positive is three true positives (0.75/0.25 = 3).
 In other words, to create a net benefit from using the model at this probability threshold, there must be more than three true positives for every false positive prediction made from the model.
@@
 
-We created a decision curve to plot net benefits across a range of probability thresholds for the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) as well as for the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
+We created a decision curve to plot net benefits across a range of probability thresholds for the aggressive alarm management strategy (alarm triggered at 15 seconds of apnea) and the conservative alarm management strategy (alarm triggered at 30 seconds of apnea).
 The decision curve takes into account the full range of reasonable clinician preferences for the point at which an alarm should be triggered to signal that the patient is apneic.
@@
 A summary of participant characteristics is presented in Table \@ref(tab:tab1).
-Nearly half of the (n=180) of the apneic events were prolonged, indicating a balanced dataset for the prediction problem at hand.
+Nearly half of the (n=180) apneic events were prolonged, indicating a balanced dataset for the prediction problem at hand.
 
@@
 
-Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution in regards to predicting durations of apneic periods.
+Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution for predicting durations of apneic periods.
 It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
@@
 For example, a recent study of the predictive ability of vital sign parameters for detection of clinical deterioration in subacute care patients yielded an AUROC score of 0.57.[@considine2020vital] When presented with a model with low AUC scores, decision curve analysis can help elicit whether the model is "good enough" to use in practice.
-Based on our results, nurses currently using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would not derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+Based on our results, nurses currently using the conservative alarm management strategy (alarm only at 30 seconds of apnea) who are willing to 'value' a false positive about 2-3 times more than a true positive would not derive an overall net benefit from using the random forest model as a trigger for apnea alarms.
 This is because using the random forest model would produce a worse outcome, in terms of the balance between true positives and false positives, for detecting if an apneic event will be prolonged, than a default strategy of waiting to trigger the alarm until the 30-second threshold.
 
-Conversely, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from instead using the random forest model as a trigger for apnea alarms.
+Conversely, nurses currently using the aggressive capnography alarm management strategy (alarm triggered at 15 seconds of apnea) who are willing to 'value' a false positive about two-thirds the importance of a true positive would derive an overall net benefit from using the random forest model as a trigger for apnea alarms.
 Using the random forest model as an additional input for an alarm trigger would reduce the total alarm burden and could be considered as an option for implementation into practice.
@@
 
-The finding that the model was able to produce an overall net benefit against the aggressive capnography alarm management strategy but not the conservative approach is a noteworthy finding.
+The finding that the model produced an overall net benefit against the aggressive capnography alarm management strategy but not the conservative approach is noteworthy.
 Further research with larger sample sizes is needed to increase the predictive power of models aimed at predicting the duration of apnea.
@@
 It would not be expected that the results of this study could be generalized to those other contexts.
+A further limitation is that, due to the observational nature of the research design, clinicians were not blinded to capnography measurements.
+It is possible that interventions implemented by clinicians during the 0 to 30-second apneic period influenced the duration of apnea.
+However, this mimics real-world practice, in that interventions may be implemented at a clinicians' discretion where no alarm conditions have been met.
+It should also be noted that only 25% of the sample had sleep apnea, which was one of the predictors included in the model.
+Due to the small sample size, the dataset used to train the model would have contained only a small portion of patients with sleep apnea and therefore it may not be generalizable to this population.
+Further research with larger sample sizes is required for confirmation.
 
@@
 The random forest model provided the best discrimination and calibration.
-Decision curve analysis indicated that using the random forest model would lead to a better outcome for capnography alarm management in comparison to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
+Decision curve analysis indicated that using the random forest model would lead to a better outcome for capnography alarm management compared to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
 The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
@@
 
-```{r tpp, fig.cap="Threshold performance plot for all models evaluated", fig.width=9}
+```{r tpp, fig.cap="Threshold performance plot for all models evaluated", fig.height=12, fig.width=15}
 targets::tar_read(tpp)
@@
 
-```{r cal, fig.cap="Calibration plot for all models evaluated"}
+```{r cal, fig.cap="Calibration plot for all models evaluated", fig.height=10}
 targets::tar_read(cal_plot)
@@
 ```{r dca, fig.cap="Decision curve analysis plots.
-Panel A is the comparison for the aggressive alarm management strategy and panel B is the comparison for the conservative alarm management strategy", fig.width=7}
+Panel A is the comparison for the aggressive alarm management strategy and panel B is the comparison for the conservative alarm management strategy", fig.width=10}
 targets::tar_read(dca_plot_combined)

### 2021-03-26 manuscript.Rmd
load data from figshare

@@
 csl: "Extras/anaesthesia.csl"
-author:
-  - Aaron Conway PhD:
-      email: aaron.conway@utoronto.ca
-      phone: +1 416 946 7112
-      address: 585 University Ave | Toronto, ON  M5G 2N2
-      institute: [PMCC, LBFON, QUT]
-      correspondence: true
-  - Carla Jungquist PhD:
-      institute: [UoB]
-      correspondence: false
-  - Kristina Chang MScN:
-      institute: [PMCC]
-      correspondence: false
-  - Navpreet Kamboj BScN:
-      institute: [LBFON]
-      correspondence: false
-  - Joanna Sutherland MBBS:
-      institute: [NSW, Coffs]
-      correspondence: false
-  - Sebastian Mafeld:
-      institute: [IR]
-      correspondence: false
-  - Matteo Parotto MD, PhD:
-      institute: [TGH, UoT]
-      correspondence: false
-institute:
-  - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
-  - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
-  - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
-  - UoB: The University of Buffalo, Buffalo, New York, USA
-  - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
-  - Coffs: Department of Anaesthesia, Coffs Harbour Health Campus, Coffs Harbour, NSW, Australia
-  - IR: Interventional Radiology, JDMI, Toronto General Hospital
-  - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
-  - UoT: Department of Anesthesiology and Pain Medicine and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
+# author:
+#   - Aaron Conway PhD:
+#       email: aaron.conway@utoronto.ca
+#       phone: +1 416 946 7112
+#       address: 585 University Ave | Toronto, ON  M5G 2N2
+#       institute: [PMCC, LBFON, QUT]
+#       correspondence: true
+#   - Carla Jungquist PhD:
+#       institute: [UoB]
+#       correspondence: false
+#   - Kristina Chang MScN:
+#       institute: [PMCC]
+#       correspondence: false
+#   - Navpreet Kamboj BScN:
+#       institute: [LBFON]
+#       correspondence: false
+#   - Joanna Sutherland MBBS:
+#       institute: [NSW, Coffs]
+#       correspondence: false
+#   - Sebastian Mafeld:
+#       institute: [IR]
+#       correspondence: false
+#   - Matteo Parotto MD, PhD:
+#       institute: [TGH, UoT]
+#       correspondence: false
+# institute:
+#   - PMCC: Peter Munk Cardiac Centre, University Health Network, Toronto, Canada
+#   - LBFON: Lawrence S. Bloomberg Faculty of Nursing, University of Toronto, Toronto, Canada
+#   - QUT: School of Nursing, Queensland University of Technology (QUT), Brisbane, Australia
+#   - UoB: The University of Buffalo, Buffalo, New York, USA
+#   - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
+#   - Coffs: Department of Anaesthesia, Coffs Harbour Health Campus, Coffs Harbour, NSW, Australia
+#   - IR: Interventional Radiology, JDMI, Toronto General Hospital
+#   - TGH: Department of Anesthesia and Pain Management, Toronto General Hospital, UHN, Toronto, Canada
+#   - UoT: Department of Anesthesiology and Pain Medicine and Interdepartmental Division of Critical Care Medicine, University of Toronto, Toronto, Canada
 output:
@@
     reference_docx: Extras/manuscript_template.docx
-    pandoc_args:
-      - --lua-filter=Extras/scholarly-metadata.lua
-      - --lua-filter=Extras/author-info-blocks.lua
+    # pandoc_args:
+    #   - --lua-filter=Extras/scholarly-metadata.lua
+    #   - --lua-filter=Extras/author-info-blocks.lua
 ---
-
-\pagebreak
-
-Declarations of interest: none
-
-Funding: A.C.
-received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S.
-Bloomberg Faculty of Nursing at the University of Toronto.
-This study received funding from the Wesley Medical Research (Project number: 2015-31).
-
-Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.
 
@@
 
-\pagebreak
-
 ## Abstract
 
-## Introduction
+### Introduction
 
@@
 
-## Methods
+### Methods
 
@@
 
-## Results
+### Results
 
@@
 
-## Conclusion
+### Conclusion
 
@@
 30 seconds).
+Preferences for the aggressive or conservative alarm threshold will be influenced by many factors, including the rate of oxygen supplmentation.
+The duration of time between the onset of apnea to hypoxemia increases with higher flows of oxygen.[@wong2019high]
 
@@
 Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.
+The raw data is archived here: https://figshare.com/account/projects/101195/articles/14320874.
 

### 2021-03-26 R/format_raw.R

@@ -2,9 +2,9 @@
 
 #' @export
 
-format_raw <- function(patient, capno_data) {
-  patient <- readRDS(patient)
-  capno_data <- readRDS(capno_data)
+format_raw <- function() {
+  patient <- readRDS(url('https://ndownloader.figshare.com/files/27274202'))
+  capno_data <- readRDS(url('https://ndownloader.figshare.com/files/27273737'))
 
   N <- length(names(capno_data))
   resp.data <- list()

### 2021-03-26 _targets.R

@@ -11,10 +11,7 @@ tar_option_set(packages = c(
 list(
   # Load in patient data and respiratory state data
   # and reformat into dataframe
-  tar_target(resp_data, format_raw(
-    patient = "data/myCapnoData.rds",
-    capno_data = "data/capno.rds"
-  )),
+  tar_target(resp_data, format_raw()),
 
   tar_target(resp_dur, as.seq(resp_data) %>%
     state.dur() %>%

### 2021-03-29 Extras/cover-letter.Rmd
submit jmir

@@
 date: "`r Sys.Date()`"
-toaddress: Journal of Radiology Nursing
+toaddress: JMIR Medical Informatics
 toname: Editorial Committee
@@
 This approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.
+
+I confirm that I agree to pay the article processing fee.

### 2021-03-29 manuscript.Rmd
submit jmir

--- 2021-03-24 manuscript.Rmd
@@
 bibliography: "Extras/references.bib"
-csl: "Extras/anaesthesia.csl"
+csl: "Extras/jmir.csl"
 author:
@@
   - Joanna Sutherland MBBS:
-      institute: [NSW, Coffs]
+      institute: [NSW]
       correspondence: false
@@
   - NSW: Rural Clinical School, University of New South Wales, Coffs Harbour, NSW, Australia
-  - Coffs: Department of Anaesthesia, Coffs Harbour Health Campus, Coffs Harbour, NSW, Australia
   - IR: Interventional Radiology, JDMI, Toronto General Hospital
@@
 
-\pagebreak
-
-Declarations of interest: none
-
-Funding: A.C.
-received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S.
-Bloomberg Faculty of Nursing at the University of Toronto.
-This study received funding from the Wesley Medical Research (Project number: 2015-31).
-
-Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.
-
 ```{r setup, echo=FALSE}
@@
 
-\pagebreak
-
 ## Abstract
 
-## Introduction
+### Background
 
 Capnography is commonly used for nurse-administered procedural sedation.
-This study aimed to determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will persist for 30 seconds or more.
-This information would help determine whether operationalizing these predictions in practice as alarm triggers would be beneficial.
+Deciphering which capnography waveform abnormalities deserve intervention, and therefore alarms to signal the event to clinicians, from those that do not is an essential step towards successfully implementing this technology into practice.
+It is possible that capnography alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
 
-## Methods
+### Objective
+
+To determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will be prolonged (defined as apnea that persisted for 30 seconds or more).
+
+### Methods
 
@@
 The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
+
+**Keywords**
+
+Procedural sedation and analgesia; conscious sedation; nursing; informatics; patient safety; machine learning; capnography; anesthesia
 
@@
 
-Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementating this technology into practice.
+Deciphering which capnography waveform abnormalities deserve intervention (and therefore alarms to signal the event to clinicians) from those that do not is an essential step towards successfully implementing this technology into practice.
 For example, triggering alarms after short periods of apnea leads to frequent interruptions and potentially increases the risk of alarm fatigue.
@@
 30 seconds).
+Preferences for the aggressive or conservative alarm threshold will be influenced by many factors, including the rate of oxygen supplmentation.
+The duration of time between the onset of apnea to hypoxemia increases with higher flows of oxygen.[@wong2019high]
 
-It is possible that capnography monitor alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
+It is possible that capnography alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
 Such an approach aligns with a call from The Society for Critical Care Medicine Alarm and Alert Fatigue Task Force, that machine learning techniques should be used to advance the quality of alerts that clinicians receive and to individualize alert delivery based on clinician response characteristics, such as alert frequency and severity.[@winters2018society]
@@
 
-Analyses were performed using R[@rcore2021].
-Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available from this webpage: <https://github.com/awconway/capnopred>.
+Analyses were performed using R version 4.0.3.[@rcore2021] Data as well as details about how to access the code and a reproducible computing environment required to verify the results is available.[@conway2021code; @conway2021data]
 
 #### Modelling
+
 We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression, and the XGBoost model.
@@
 Pre-processing steps included normalizing numeric predictors and using an interaction term for the duration of the previous respiratory state and the total number of apneic events.
+Discriminative ability of the models was compared using the area under receiver operating characteristics curve (AUROC) as well as by plotting sensitivity, specificity, positive predictive values and negative predictive values (termed a threshold performance plot).
+A calibration plot with a loess smoother was used to assess calibration.[@van2016calibration] The `runway` package was used to create the plots.[@singh2021]
 
@@
 A summary of participant characteristics is presented in Table \@ref(tab:tab1).
-Nearly half of the (n=180) apneic events were prolonged, indicating a balanced dataset for the prediction problem at hand.
+Nearly half of the (n=180) apneic events were prolonged (i.e.
+>30 seconds), indicating a balanced dataset for the prediction problem at hand.
 
@@
 
-The random forest had the best discriminatory power of the models, with a mean AUROC score of 0.66 (SE 0.03).
+The random forest had the best discriminatory power of the models, with a mean AUROC score of 0.66 (standard error 0.03).
 A threshold performance plot, which summarises the discriminatory power for the models, including values for sensitivity, specificity, positive predictive value, and negative predictive value across all probability thresholds, is presented in Figure \@ref(fig:tpp).
@@
 Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution for predicting durations of apneic periods.
-It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (HR 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (HR 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
+It was found in a study of capnography waveform abnormalities during nurse-administered sedation that the hazard for apnea increased two-fold (hazard ratio 2.14, 95% CI 1.75 - 2.62) when a patient was in a state of hypoventilation (defined as >10% reduction in ETCO~2~ from baseline).[@conway2019pre] The risk of apnea also increased with each additional sedation dose (hazard ratio 2.86; 95% CI 2.15 - 3.81).[@conway2019pre] Results from an earlier study in a different population also supported the observations that the onset of apneic periods during sedation are associated with the history of previous respiratory states.
 Krauss and colleagues[@krauss2016] used survival analysis to model the time to first apneic events in a sample of 312 patients undergoing procedural sedation with propofol or ketamine in the emergency department.
@@
 Using the random forest model as an additional input for an alarm trigger would reduce the total alarm burden and could be considered as an option for implementation into practice.
-To operationalize these predictions into capnography monitors, partnerships with industry would be required.[@glasgow2018nurse] Integrating predictive models into alarm management strategies for respiratory monitoring devices is also indicated in other contexts.
+To operationalize these predictions into capnography monitors, partnerships with industry would be required because monitor functionality would need to be adapted to facilitate input of the data required to calculate the predictions.[@glasgow2018nurse] These would include patient characteristics and sedation dosing.
+Integrating predictive models into alarm management strategies for respiratory monitoring devices is also indicated in other contexts.
 For example, a recent study found that opioid-induced respiratory depression during recovery from anesthesia can be accurately predicted using a machine learning approach.[@jungquist2019identifying] In addition, user-centered design considerations, such as how the predictions should be communicated to nurses responsible for decision-making, would be important avenues for further research prior to implementation.[@risling2020advancing]
@@
 Yet, previous research indicates that finding such a solution may be worthwhile.
-An analysis of over 500 thousand patients found that respiratory compromise in interventional radiology procedures performed with moderate sedation contributed to worse clinical outcomes and higher costs.[@urman2019impact]
+An analysis of half a million patients found that respiratory compromise in interventional radiology procedures performed with moderate sedation contributed to worse clinical outcomes and higher costs.[@urman2019impact]
 
@@
 The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
+
+\pagebreak
+
+## Ackowledgements
+
+A.C.
+received support from a National Health and Medical Research Council Early Career Fellowship (APP1091657) and the Bertha Rosenstadt Faculty Small Research Grant from the Lawrence S.
+Bloomberg Faculty of Nursing at the University of Toronto.
+This study received funding from the Wesley Medical Research (Project number: 2015-31).
+Role of the funding source: The funders had no role in study design; in the collection, analysis and interpretation of data; in the writing of the report; and in the decision to submit the article for publication.
+
+## Conflicts of interest
+
+None
+
+## Abbreviations
+
+AUROC: Area under the receiver operating characteristics curve
+ETCO~2~: End-tidal carbon dioxide
+CO~2~: Carbon dioxide
 

### 2021-05-26 manuscript.Rmd
R1 submitted

@@
 ---
-title: "Predicting prolonged apnea during nurse-administered procedural sedation using machine learning"
+title: "Predicting prolonged apnea during nurse-administered procedural sedation: A machine learning study"
 bibliography: "Extras/references.bib"
@@
 Results for the primary aim of the observational study are reported elsewhere.[@conway2019sequence] All participants provided written informed consent and the study was approved by human research ethics committees (UCH HREC 1614; SVHAC HREC 16/26; QUT 1600000641).
-The study was undertaken in accordance with the Australian National Statement on Ethical Conduct in Human Research[@national2007] and it was registered prospectively (ACTRN12616001132437).
 
@@
 
-Demographic data and clinical characteristics were collected from medical records and self-report questionnaires prior to procedures.
+Data were collected from August 2016 to May 2018.
+Demographic data and clinical characteristics were collected from medical records or directly from participants prior to procedures.
 Intra-procedural data were collected in real-time by the researcher who was present in the procedure room.
@@
 Nearly half of the (n=180) apneic events were prolonged (i.e.
->30 seconds), indicating a balanced dataset for the prediction problem at hand.
+>30 seconds).
+
+```{r tab.cap='Participant characteristics', tab.id='tab1'}
+targets::tar_read(summary_table)
+```
 
@@
 
+In this study, we found that a random forest model had the best discriminative ability and calibration for predicting if an apneic event during nurse-administered procedural sedation would be prolonged.
+However, it should be noted that the accuracy of this random forest model was still quite low (AUROC = 0.66).
+Additional research should be undertaken with larger sample sizes to validate our initially promising findings.
+
 Results from prior research indicated that using information about the history of previous respiratory states may be a promising solution for predicting durations of apneic periods.
@@
 It was identified that having an abnormal ETCO~2~ measurement 30, 60 and 90 seconds prior to an apneic event increased risk for apnea (hazard ratios with 95% CI for the different lag durations of 2.45 (1.63 - 3.69), 1.88 (1.21 - 2.92) and 2.06 (1.36 - 3.11), respectively).
-
-We leveraged these preliminary findings about the associations between apneic events and the history of previous respiratory states in the present study by building a predictive model using a machine learning approach.
-A random forest model had the best discriminative ability and calibration from those that we tested.
-However, it should be noted that the accuracy of this random forest model was still quite low (AUROC = 0.66).
-Additional research should be undertaken with larger sample sizes to validate our initially promising findings.
+In our study, ee leveraged information about the associations between apneic events and the history of previous respiratory states by building a predictive model using a machine learning approach.
+Features included in the models we tested were the previous respiratory state, duration of time in the previous respiratory state, number of previous apneic events and the duration of the previous apneic event.
 
@@
 For example, a recent study found that opioid-induced respiratory depression during recovery from anesthesia can be accurately predicted using a machine learning approach.[@jungquist2019identifying] In addition, user-centered design considerations, such as how the predictions should be communicated to nurses responsible for decision-making, would be important avenues for further research prior to implementation.[@risling2020advancing]
+
+This study used decision curve analysis as a way to evaluate the potential clinical impact that using the model as input for capnography alarm management would have on the number of alarms triggered (i.e.
+false positives/negatives).
+However, as with any intervention in healthcare, efficacy needs to be assessed prior to broader implementation.
+Efficacy in this context would be whether using the model as input for capnography alarm management improves patient safety.
+The gold standard approach for such an evaluation is a randomized controlled trial.
+Randomized controlled trials testing alarm conditions that have integrated predictions from machine learning models have been conducted previously in similar contexts, such as intra-operative blood pressure management.[@maheshwari2020hypotension; @wijnberge2020effect]
 
@@
 It should also be noted that the data used in this analysis were produced from an observational study conducted at two hospitals that used a convenience sampling approach.
-Therefore the possibility of selection bias should be considered.
+Therefore, the possibility of selection bias should be considered.
 The context in which the study was conducted should also be considered in regard to external validity.
@@
 
-## Ackowledgements
+## Acknowledgements
 
@@
 
-```{r tab.cap='Participant characteristics', tab.id = 'tab1'}
-targets::tar_read(summary_table)
-```
-
-\pagebreak
-
 ```{r roc, fig.cap="Area under receiver operating characteristics curve"}

### 2021-05-26 _targets.R

@@ -1,12 +1,11 @@
 library(targets)
 library(tarchetypes)
 
-devtools::load_all()
 options(tidyverse.quiet = TRUE)
 
 tar_option_set(packages = c(
   "tidymodels", "tidyverse",
-  "patchwork"
+  "patchwork", "capnopred"
 ))
 list(
   # Load in patient data and respiratory state data
@@ -31,7 +30,7 @@ list(
   tar_target(data_split, combined %>%
     make_train_id(
       seed = 42,
-      frac = 0.75
+      frac = 0.75 
     )),
 
   tar_target(

### 2021-12-09 manuscript.Rmd
minor figure changes for revision

@@
 
-### Background
+### Aim
 
-Capnography is commonly used for nurse-administered procedural sedation.
-Deciphering which capnography waveform abnormalities deserve intervention, and therefore alarms to signal the event to clinicians, from those that do not is an essential step towards successfully implementing this technology into practice.
-It is possible that capnography alarm management may be improved by using machine learning to create a ‘smart alarm’ that can alert clinicians for apneic events that are predicted to be prolonged.
+Determine the accuracy of machine learning models for predicting if apnea during nurse-administered procedural sedation will be prolonged (>30 seconds).
 
-### Objective
+### Design
 
-To determine the accuracy of machine learning models for predicting, at the 15-second time point, if apnea will be prolonged (defined as apnea that persisted for 30 seconds or more).
+Secondary analysis of an observational study.
 
@@
 
-A secondary analysis of an observational study was conducted.
-We selected several candidate models to evaluate, including a random forest model, generalized linear model (logistic regression), lasso regression, ridge regression and the XGBoost model.
-Out of sample accuracy of the models was calculated using 10-fold cross-validation.
-The net benefit decision analytic measure was used to assist with deciding whether using the models in practice would lead to better outcomes on average than the default capnography alarm management strategies currently in place.
-The default strategies are: 1) the aggressive approach, which involves triggering an alarm after brief periods of apnea (typically 15 seconds); and 2) the conservative approach, which involves triggering an alarm for only prolonged periods of apnea (typically 30 seconds).
+A selection of candidate models were evaluated.
+Out of sample accuracy was calculated using 10-fold cross-validation.
+Decision analysis was used to decide if the models were better, in terms of false positive and negatives, than alternative alarm management strategies, being: an aggressive approach (alarm after 15 seconds); or a conservative approach (alarm after 30 seconds).
 
@@
 
-A total of 384 apneas longer than 15 seconds were observed in 61 of the 102 patients who participated in the observational study.
+A total of 384 apneas longer than 15 seconds were observed in 61 of the 102 participants.
 Nearly half of the apneas (n=180) were prolonged.
 The random forest had the best discrimination (AUROC 0.66) and calibration.
-Net benefit associated with the random forest model exceeded the aggressive alarm management strategy but was lower than the conservative alarm management strategy.
-
-### Conclusion
-
-Decision curve analysis indicated that using a random forest model would lead to a better outcome for capnography alarm management compared to an aggressive strategy where alarms are triggered after 15 seconds of apnea.
-The model would not be superior to the conservative strategy, where alarms are only triggered after 30 seconds.
+Net benefit associated with a random forest model exceeded the aggressive alarm management strategy but was lower than the conservative alarm management strategy.
 
@@
 For this reason, implementing capnography into practice for respiratory monitoring is considered a high priority to improve patient safety by leading authorities, including national and international professional organizations for anesthesiology in Canada, the United States, and Europe.[@apfelbaum2018; @hinkelbeinEuropeanSocietyAnaesthesiology2018; @Dobson2018] It is commonly used for nurse-administered procedural sedation, [@Conway_2014; @conway2013; @conway2013a] including in the interventional radiology setting.[@pella2018systematic; @brast2016capnography; @long2016capnography; @mafeld2020avoiding; @werthman2020administration]
+
+## Background
 
@@
 The primary aim of the observational study was to identify common patterns in capnography waveform abnormalities and factors that influence these patterns.
-Results for the primary aim of the observational study are reported elsewhere.[@conway2019sequence] All participants provided written informed consent and the study was approved by human research ethics committees (UCH HREC 1614; SVHAC HREC 16/26; QUT 1600000641).
+Results for the primary aim of the observational study are reported elsewhere.[@conway2019sequence]
+
+### Ethics
+
+All participants provided written informed consent and the study was approved by human research ethics committees (UCH HREC 1614; SVHAC HREC 16/26; QUT 1600000641).
 
@@
 A total of 18 predictor variables were used.
+No imputation of missing data was performed.
+No imputation of missing data was performed.
 

### 2021-12-09 R/cal_plots.R

@@ -1,0 +1,204 @@
+#' Generate a single calibration plot with error bars showing 95 percent
+#' confidence intervals
+#'
+#' This code builds off of code written by Darren Dahly, PhD in this blog post:
+#' \url{https://darrendahly.github.io/post/homr/}.
+#'
+#' @param df The df as a data.frame.
+#' @param outcome A character string containing the name of the column
+#'   containing the outcomes (expressed as 0/1s).
+#' @param prediction A character string containing the name of the column
+#'   containing the predictions.
+#' @param n_bins Number of bins. Defaults to 10. Set to 0 to hide binned
+#'   calibration.
+#' @param show_loess Whether to show loess smoothed calibration estimates.
+#'   Defaults to FALSE. For \code{cal_plot()}, you can display both a binned
+#'   calibration plot and a loess curve. In contrast, for
+#'   \code{cal_plot_multi()}, these options are mutually exclusive.
+#' @param plot_title A character string containing the title for the resulting
+#'   plot.
+#' @return A ggplot containing the calibration plot
+#' @examples
+#' data(single_model_dataset)
+#' cal_plot(single_model_dataset, outcome = 'outcomes', prediction = 'predictions', n_bins = 5)
+#' @export
+cal_plot <- function(df, outcome, prediction, n_bins = 10, show_loess = FALSE, plot_title = '', ...){
+
+  # The calibration plot
+  g1 <- dplyr::mutate(df, bin = dplyr::ntile(!!rlang::parse_expr(prediction), n_bins)) %>%
+    # Bin prediction into n_bins
+    dplyr::group_by(bin) %>%
+    dplyr::mutate(n = dplyr::n(), # Get ests and CIs
+           bin_pred = mean(!!rlang::parse_expr(prediction), na.rm = TRUE),
+           bin_prob = mean(as.numeric(as.character(!!rlang::parse_expr(outcome))), na.rm = TRUE),
+           se = sqrt((bin_prob * (1 - bin_prob)) / n),
+           ul = bin_prob + 1.96 * se,
+           ll = bin_prob - 1.96 * se) %>%
+    dplyr::mutate_at(dplyr::vars(ul, ll), . %>% scales::oob_squish(range = c(0,1))) %>%
+    dplyr::ungroup() %>%
+    ggplot2::ggplot() +
[…]

### 2021-12-09 R/combine_preds.R

@@ -10,7 +10,7 @@ combine_preds <- function(rf_preds,
   rf_preds$model_name <- "Random forest"
   boost_preds$model_name <- "XGboost"
   glm_preds$model_name <- "Logistic regression"
-  lasso_preds$model_name <- "Lasso regression"
+  lasso_preds$model_name <- "LASSO regression"
   ridge_preds$model_name <- "Ridge regression"
 
   bind_rows(

### 2021-12-09 _targets.R

@@ -125,13 +125,14 @@ list(
 
   tar_target(
     cal_plot,
-    runway::cal_plot_multi(combined_preds,
+    cal_plot_multi(combined_preds,
       outcome = "outcome",
       prediction = ".pred_long",
       show_loess = TRUE,
       n_bins = 0,
       model = "model_name"
     )
+    
   ),
   tar_target(
     roc_plot,
@@ -236,7 +237,7 @@ list(
     select(-`Random forest_sm`,
       "Threshold" = threshold,
       -all,
-      "Alarm at 30 seconds" = none
+      "Alarm at 30 s" = none
     ) %>%
     pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
     ggplot(aes(x = Threshold, y = `Net benefit`)) +
@@ -259,7 +260,7 @@ list(
   tar_target(dca_plot15, dca15$net.benefit %>%
     select(-`Random forest_sm`,
       "Threshold" = threshold,
-      "Alarm at 15 seconds" = all,
+      "Alarm at 15 s" = all,
       -none
     ) %>%
     pivot_longer(cols = -Threshold, names_to = "Decisions", values_to = "Net benefit") %>%
